R for Clinical Study Reports and Submission

### 2021-11-15 tlf-disposition.Rmd
Xiaoxia reviewed TLFs part on Nov14

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
a summary needs to be provided to account for all participants who entered the study in Section 10.1, Disposition of Participants.

The disposition of participant reports the numbers of participants who were randomized,
and who entered and completed each phase of the study.
In addition, the reasons for all post-randomization discontinuations,
grouped by treatment and by major reason (lost to follow-up, adverse event, poor compliance, etc.).

@@

In this chapter, the creation of a typical disposition table is shown.

@@

The first step is to read in the relevant datasets into R.
For a disposition table, all the required information is saved in a Subject-level Analysis Dataset (ADSL).
The dataset is provided in `sas7bdat` format that is a SAS data format currently used in many clinical trial analysis and reporting.
The `haven` package is able to read in the dataset, while maintaining its attributes (e.g., variable labels).

@@

Following variables are used in the preparation of a simplified disposition of participants table:

- USUBJID: Unique subject identifier
- TRT01P: Planned treatment
- TRT01PN: Planned treatment numeric encoding
- DISCONFL: Discontinued from study flag
- DCREASCD: Discontinued from study reason coded

@@

In the code below, we calculated participants in the analysis population.

@@

In the code below, we calculate the number and percentage of
participants who completed/discontinued the study for different reasons.

@@

In the code below, we calculate the number and percentage of participants who complete the study.
We split `n_reason` because we need to customize the row order of the table.

@@

In the code below, we calculate the number and percentage of participants who discontinued the study for different reasons.
For display purposes, `paste0("    ", row)` is used to add leading spaces to mimick indentation in the final report.

@@

Now we combined individual rows into the whole table for reporting purpose.
`tbl_disp` is used as input for `r2rtf` to create final report.

@@

- Step 1: Read subject level data into R, i.e., `adsl`.
- Step 2: Count participants in the analysis population and name the dataset `n_rand`.
- Step 3: Count the number and percentage of participants that discontinued the study, and name the dataset `n_disc`.
- Step 4: Count the number and percentage of participants that discontinued the study for different reasons,
  and name the dataset `n_reason`.
- Step 5: Count the number and percentage of participants that completed the study, and name the dataset `n_complete`.
- Step 6: Bind `n_rand`, `n_disc`, `n_reason`, and `n_complete` by row.
- Step 7: Write the final table to RTF

>>>>

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
a summary table needs to be provided to include all participants who entered the study in Section 10.1, Disposition of Participants.

The disposition of participants table reports the numbers of participants who were randomized,
and who entered and completed each phase of the study.
In addition, the reasons for all post-randomization discontinuations,
grouped by treatment and by major reason (lost to follow-up, adverse event, poor compliance, etc.) are reported.

@@

In this chapter, we show how to the create a typical disposition table.

@@

The first step is to read in the relevant datasets into R.
For a disposition table, all the required information is saved in a Subject-level Analysis Dataset (ADSL).
This dataset is provided in `sas7bdat` format, which is a SAS data format currently used in many clinical trial analysis and report.
The `haven` package is able to read the dataset, while maintaining its attributes (e.g., variable labels).

@@

The following variables are used in the preparation of a simplified disposition of participants table:

- USUBJID: unique subject identifier
- TRT01P: planned treatment
- TRT01PN: planned treatment numeric encoding
- DISCONFL: discontinued from study flag
- DCREASCD: discontinued from study reason coded

@@

In the code below, we calculate the number of participants in the analysis population by treatment arms.

@@

In the code below, we calculate the number and percentage of
participants who completed/discontinued the study for different reasons by treatment arms.

@@

In the code below, we calculate the number and percentage of participants who complete the study by treatment arms.
We split `n_reason` because we want to customize the row order of the table.

@@

In the code below, we calculate the numbers and percentages of participants who discontinued the study for different reasons by treatment arms.
For display purpose, `paste0("    ", row)` is used to add leading spaces to produce indentation in the final report.

@@

Now we combine individual rows into one table for reporting purpose.
`tbl_disp` is used as input for `r2rtf` to create final report.

@@

- Step 1: Read subject level data (i.e., `adsl`) into R.
- Step 2: Count participants in the analysis population and name the dataset `n_rand`.
- Step 3: Calculate the number and percentage of participants who discontinued the study by treatment arm, and name the dataset `n_disc`.
- Step 4: Calculate the numbers and percentages of participants who discontinued the study for different reasons by treatment arm,
  and name the dataset `n_reason`.
- Step 5: Calculate the number and percentage of participants who completed the study by treatment arm, and name the dataset `n_complete`.
- Step 6: Bind `n_rand`, `n_disc`, `n_reason`, and `n_complete` by row.
- Step 7: Write the final table to RTF
